---
title: 'Reproducible Research: Peer Assessment 2'
author: "Aneesh Sathe"
date: "June 12, 2015"
output: html_document
---

#Synopsis:
at most 10 complete sentences

#Data Processing

```{r Read Data, cache=TRUE}
library(data.table)
in_dat=data.table(read.table("repdata-data-StormData.csv.bz2",sep = ",",header = TRUE))

```

```{r}
in_dat$BGN_DATE=as.Date(in_dat$BGN_DATE)
```

Your data analysis must address the following questions:

Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Across the United States, which types of events have the greatest economic consequences?

#Results


at least one figure containing a plot

no more than three figures

Publishing Your Analysis

For this assignment you will need to publish your analysis on RPubs.com. If you do not already have an account, then you will have to create a new account. After you have completed writing your analysis in RStudio, you can publish it to RPubs by doing the following:

In RStudio, make sure your R Markdown document (.Rmd) document is loaded in the editor

Click the Knit HTML button in the doc toolbar to preview your document.

In the preview window, click the Publish button.

Once your document is published to RPubs, you should get a unique URL to that document. Make a note of this URL as you will need it to submit your assignment.

NOTE: If you are having trouble connecting with RPubs due to proxy-related or other issues, you can upload your final analysis document file as a PDF to Coursera instead.